Citation Analysis of Computer Science

### 2019-11-21 Scientometrics/comp.tex
added a new figure

Computer science, and its applications, has experienced rapid growth and diversification over the last twenty years. The collective influence of the Internet of Things (IoT), `big' data, accessible cloud computing, and advances in artificial intelligence have further been postulated as a recent driver for its growth and evolution~\cite{siebel2019_digital}. As also noted in a 2017 US National Academies Report~\cite{nas_2017}, ``A wide range of jobs in virtually all sectors demand computing skills to an unprecedented extent. And every academic discipline finds itself incorporating computing into its research and educational mission". Given this powerful influence of computer science in research, some understanding of its present state, structure, and relationship to other fields that seems warranted since it can inform planning and policy making at multiple levels from national level funding all the way down to faculty hiring strategy. For the same reason of its broad influence, computer science can also be an excellent model system for studying interdisciplinarity. These are not novel ideas, however, as far back as 1979, Salton and Bergmark conducted a historical study of the computer science literature (419 computer articles published in 1974,  and 3,812 references cited in these articles) in which they made the key point that the scientific literature serves a rich source of information to study the structure and historical development of a field~\cite{salton_citation_1979}. These authors described the global structure of computer science as comprising three main subareas (i) theoretical foundations, such as theory of computation, (ii) hardware and computer systems, such as architecture,  and (iii) software, such as programming systems.  Related areas acknowledged were  (a) mathematics of computing, such as numerical analysis, (b) special software topics, such as operating systems, (c) data management and database systems, (d) methodologies valid for multiple applications, such as algebraic manipulation, (e) computer applications, such as computer graphics, (f) non-technical aspects, such as computer education.

Much in the way of growth and evolution of computing has occurred since 1974. Progressing beyond the historical triad of theoretical foundations; hardware and computer systems; and software, the Computing Classification System (CCS)~\cite{acm_ref} published by the Association for Computing Machinery now consists of 13 top-level areas that reflect a more current view of the field. This classification does address relationships with other fields under Applied Computing. However, an easy way to map scientific publications, especially interdisciplinary articles or those from adjacent fields, to the CCS classification does not exist.

Other classification systems are available such as the All Science Journal Classification (ASJC) developed and maintained by Scopus. the Web of Science research and categorical classifications, the NSF classification system~\cite{nsf_classification}. All three rely on applying one or more journal-derived labels to articles. Disciplinary assignments using journal based classifications have been extensively discussed and critiqued~\cite{wang_large-scale_2016} [need comprehensive list of citations] with Shu and colleagues recently noting from a comparative study of the Chinese Science Citation Database (CSCD) and the Web of Science that 46\% of articles did not belong to the discipline of the journal they were published in~\cite{shu_comparing_2019}. Article classification systems have been developed, however~\cite{traag_louvain_2019,boyack_classification_2014,waltman_new_2012} that escape some of the criticisms of journal-based classification.

@@

In extending the work of Salton and Bergmark to the 21st century, we sought to combine journal and article based classifications in a scalable manner. Consequently, we used a whole text search method to match records from two major data sources: (i) the DBLP bibliography--a reference bibliography for computer science~\cite{dblp_ref} and  (ii) the journal-based Scopus bibliography, which more comprehensively covers the physical sciences, life sciences, social sciences, and health sciences. 
\textcolor{blue}{Do we need to cite Scopus?}
The underlying assumption is that a record in DBLP qualifies as computer science. Therefore, we cross-matched DBLP publications to Scopus to  enable identification of linked publications across multiple disciplines.

Using both direct citations and co-citations as the basis for clustering, we used articles from from journals and conference proceedings for the period 1996-2015. We reconciled these clusters with the the All Science Journal Classification (ASJC) developed and maintained by Scopus as well as to the CCS classification through a combination of automated and manual procedures. We describe below the high-level interactions of computer science within the Physical Sciences as well as with the Social Sciences and Life Sciences.

@@

\subsection{Data}
 A stable release of the DBLP computer science bibliography~\cite{dblp_ref} consisting of 7,079,994 records was downloaded as dblp-2018-08-01.xml.gz. Publications were parsed from the xml source file and loaded into a PostgreSQL database. We have previously parsed, the Scopus dataset of over 72,000,000 publications into a custom schema as part of a larger data platform for research evaluation~\cite{GithubERNIE2019}. Records in the DBLP dataset, were matched to Scopus identifiers using PostgreSQL's full text search feature applied to document titles. The process involves parsing all Scopus document titles into tokens, reducing the tokens into lexemes and returning vectors containing these lexemes. The vectors are then matched with corresponding text search vectors extracted from DBLP document titles. A rank is returned based on the frequency of matching lexemes. Only records in DBLP with DOIs (4,291,130) were considered and only matches with a normalized rank of greater than 0.99 were accepted, roughly equivalent to `at least 11 matching lexemes independent of order and vector length'. This procedure resulted in a dataset of 634,100 DBLP publications with Scopus identifiers. Cited references were extracted from Scopus, resulting in a total of 3,960,503 nodes (publications and references) and 11,992,387 edges that represent citations of these references and publications within the dataset. This dataset is referred to as \emph{comp}.\par

@@

\paragraph{Clustering by Co-Citation} To build clusters using co-citations, we used variable level clustering, an approach developed by Small and Sweeney (1985)~\cite{small_clustering_1985}, with minor modifications. Variable level clustering involves applying a threshold (below which all edges are dissolved) then iteratively selecting edges with the highest normalized co-citation value and extracting connected components from the graph as clusters for each edge in turn. Three parameters are needed (i) a threshold or starting level based on a quantile of normalized co-citation frequency (ii) a level increment (iii) a maximum cluster size. An issue is the generation of very large clusters by chaining through low edge weights.  Thus, at each iteration, any cluster exceeding the maximum cluster size is put back into the process and a higher threshold applied to break such clusters.

@@

We applied variable clustering with initial parameters of threshold ($t$) = 0.5 (normalized co-citation frequency, see Methods), increment $i = 0.1$, and maximum cluster size, $mcs = 100$. Thus, at start, all edges below the median normalized co-citation frequency are dissolved. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5-0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster. All clusters generated thus, contained  less than 100 nodes.  Using this approach, 6,822 clusters containing 35,485 nodes were generated. Clusters containing only 2 nodes were then discarded bringing the total number of clusters down to 3,851. Pairwise clustering was then performed on these clusters to generate super-clusters. After 9 rounds of pairwise clustering, 13 clusters were generated resulting from agglomeration of 3,851 clusters

@@

For the purpose of this study, we defined, the computer science literature as a working set of all articles in the DBLP bibliography published in 1996-2015, that had a digital object identifier (DOI), and could be matched to article identifiers in the Scopus bibliography. The process of selection and matching resulted in a dataset of 634,100 publications (Materials and Methods). We constructed article level clusters of the computer science literature for the years 1996-2015, at a sufficiently high level of aggregation to be comparable to disciplinary classification by journal. Clustering of publications is commonly accomplished through direct citation, bibliographic coupling, and co-citation with an argument being made for direct citation as the best approach to concentrate citation links~\cite{klavans_which_2017}. Accordingly, we first used direct citation links as the basis for cluster formation. A set of 20 clusters was generated using a spectral clustering technique, Graclus, that has previously been applied to citation data~\cite{graclus_2007,subelj_clustering_2016}. To place these clusters in context, we identified ASJC categories that accounted for at least 10\% of the publications in each cluster and observed 31 of the 334 minor subject areas in the Scopus ASJC classification. (which consists of 5 subject areas, 27 major subject areas, and 334 minor subject areas).

@@

Here we describe the overall strategy used for the two clustering methods.\\
\begin{figure}[ht]
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.5]{scopus_dblp_graclus2.pdf}
% figure caption is below the figure
\caption{caption will be inserted I just haven't been able to overcome the threshold to avoid boring tasks}
\label{heatmap}       % Give a unique label
\end{figure}

>>>>

Computer science, and its applications, has experienced rapid growth and diversification over the last twenty years. The collective influence of the Internet of Things (IoT), `big' data, accessible cloud computing, and advances in artificial intelligence have further been postulated as a recent driver for its growth and evolution~\cite{siebel2019_digital}. As also noted in a 2017 US National Academies Report~\cite{nas_2017}, ``A wide range of jobs in virtually all sectors demand computing skills to an unprecedented extent. And every academic discipline finds itself incorporating computing into its research and educational mission". Given this powerful influence of computer science in research, some understanding of its present state, structure, and relationship to other fields that seems warranted since it can inform planning and policy making at multiple levels from national level funding all the way down to faculty hiring strategy. For the same reason of its broad influence, computer science can also be an excellent model system for studying interdisciplinarity. These are not novel ideas, however, as far back as 1979, Salton and Bergmark conducted a historical study of the computer science literature (419 computer articles published in 1974,  and 3,812 references cited in these articles) in which they made the key point that the scientific literature serves a rich source of information to study the structure and historical development of a field~\cite{salton_citation_1979}. These authors described the global structure of computer science as comprising three main subareas (i) theoretical foundations, such as theory of computation, (ii) hardware and computer systems, such as architecture,  and (iii) software, such as programming systems.  Related areas described were  (a) mathematics of computing, such as numerical analysis, (b) special software topics, such as operating systems, (c) data management and database systems, (d) methodologies valid for multiple applications, such as algebraic manipulation, (e) computer applications, such as computer graphics, (f) non-technical aspects, such as computer education.

Computing has grown and evolved significantly since 1974. Progressing beyond the historical triad of theoretical foundations; hardware and computer systems; and software\cite{salton_citation_1979}, the Computing Classification System (CCS)~\cite{acm_ref} published by the Association for Computing Machinery now consists of 13 top-level areas that reflect a more current view of the field. This classification does address relationships with other fields under Applied Computing. However, an easy way to map scientific publications, especially interdisciplinary articles or those from adjacent fields, to the CCS classification does not exist.

Other classification systems are available such as the All Science Journal Classification (ASJC) developed and maintained by Scopus. the Web of Science research and categorical classifications, the NSF classification system~\cite{nsf_classification}. Of these three, Scopus and the Web of Science make also contain citation data and proprietary unique identifiers for publications that can be matched, in some cases, to digital object identifiers (DOIs). All three rely on applying one or more journal-derived labels to articles. Shu and colleagues recently noted in a comparative study of the Chinese Science Citation Database (CSCD) and the Web of Science that 46\% of articles did not belong to the discipline of the journal they were published in~\cite{shu_comparing_2019}. Others have also discussed and critiqued disciplinary assignments using journal based classifications~\cite{wang_large-scale_2016} \textcolor{blue}{need comprehensive list of citations}. Article classification systems have been explored~\cite{traag_louvain_2019,boyack_classification_2014,waltman_new_2012}, which escape some of the criticisms of journal-based classification.

@@

In extending the work of Salton and Bergmark to the 21st century, we were also interested in connections between computer science and other fields. As a source of computer science literature, we used DBLP, a reference bibliography for computer science~\cite{dblp_ref}. The DBLP bibliography covers publications from computer science and includes publications from hybrid fields in cases they are considered pertinent to computer science research. The underlying assumption is that records in DBLP are greatly enriched for computer science even while recognizing that coverage is incomplete and possibly biased. Cross-matching DBLP publications to Scopus records using whole text searches applied to titles, enabled harvesting citation data as well as publications from other disciplines. This cross-matching also allows the use of both journal and article based classifications when clustering documents. Using both direct citations and co-citations as the basis for clustering, we used articles from from journals and conference proceedings for the period 1996-2015. We reconciled these clusters with the the All Science Journal Classification (ASJC) developed and maintained by Scopus as well as to the CCS classification through a combination of automated and manual procedures. We describe below the high-level interactions of computer science within the Physical Sciences as well as with the Social Sciences and Life Sciences.

@@

\subsection{Data}
 A stable release of the DBLP computer science bibliography~\cite{dblp_ref} consisting of 7,079,994 records was downloaded as dblp-2018-08-01.xml.gz. Publications were parsed from the xml source file and loaded into a PostgreSQL database. We have previously parsed, the Scopus dataset of over 72,000,000 publications into a custom schema as part of a larger data platform for research evaluation~\cite{GithubERNIE2019}. Records in the DBLP dataset, were matched to Scopus identifiers using PostgreSQL's full text search feature applied to document titles. The process involves parsing all Scopus document titles into tokens, reducing the tokens into lexemes and returning vectors containing these lexemes. The vectors are then matched with corresponding text search vectors extracted from DBLP document titles. A rank is returned based on the frequency of matching lexemes. Only records in DBLP with DOIs (4,291,130) were considered and only matches with a normalized rank of greater than 0.99 were accepted, roughly equivalent to `at least 11 matching lexemes independent of order and vector length'. This procedure resulted in a dataset of 634,100 DBLP publications with Scopus identifiers where 49.6\% were labeled as article and 48.8\% as conference proceeding. Cited references were extracted from Scopus, resulting in a total of 3,960,503 nodes (publications and references) and 11,992,387 edges that represent citations within the dataset. This dataset is referred to as \emph{comp}.\par

 \begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.5]{ar_cp_ratio.pdf}
% figure caption is below the figure
\caption{Summary of DBLP data cross-matched with Scopus. 634,100 publications from DBLP were cross-matched with Scopus and then matched to the 27 major subject areas in the ASJC (Scopus) classifications. The largest number of publications are contributed by Computer Science; Mathematics; and Engineering and then by Social Sciences; Physics and Astronomy; and Biochemistry, Genetics, and Molecular Biology. Publications were further annotated with respect to being either articles or conference proceedings and ordered on the abscissa by the ratio of conference proceedings to articles within a major subject area (cp\_count/ar\_count). For this dataset, the major subject areas of Energy; Computer Science; Medicine; Mathematics; and Engineering had the highest ratio of conference proceedings to articles while Immunology and Microbiology; Chemistry; General (multidisciplinary); Pharmacology, Toxicology, and Pharmaceutics; and Agricultural and Biological Sciences had the lowest ratio. }
\label{ar_cp_annotation}       % Give a unique label
\end{figure}

@@

\paragraph{Clustering by Co-Citation} To build clusters using co-citations, we used variable level clustering, an approach developed by Small and Sweeney (1985)~\cite{small_clustering_1985}, with minor modifications. Variable level clustering involves applying a threshold (below which all edges are dissolved) then iteratively selecting edges with the highest normalized co-citation value and extracting connected components from the graph as clusters for each edge in turn. Three parameters are needed (i) a threshold or starting level based on a quantile of normalized co-citation frequency (ii) a level increment (iii) a maximum cluster size. An issue is the generation of very large clusters by chaining through low edge weights.  Thus, at each iteration, any cluster exceeding the maximum cluster size returned to the process and a higher threshold applied to break such clusters.

@@

We applied variable clustering with initial parameters of threshold ($t$) = 0.5 (normalized co-citation frequency, see Methods), increment $i = 0.1$, and maximum cluster size, $mcs = 100$. Thus, at start, all edges below the median normalized co-citation frequency are dissolved. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5-0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster. All clusters generated thus, contained  less than 100 nodes.  Using this approach, 6,822 clusters containing 35,485 nodes were generated. Clusters containing only 2 nodes were then discarded bringing the total number of clusters down to 3,851. Pairwise clustering was then performed on these clusters to generate higher-order clusters. After 9 rounds of pairwise clustering, 13 clusters were generated resulting from agglomeration of 3,851 clusters

@@

\begin{figure}[ht]
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.5]{scopus_dblp_graclus2.pdf}
% figure caption is below the figure
\caption{caption will be inserted I just haven't been able to overcome the threshold to avoid boring tasks}
\label{heatmap}       % Give a unique label
\end{figure}
For the purpose of this study, we defined, the computer science literature as a working set of all articles in the DBLP bibliography published in 1996-2015, that had a digital object identifier (DOI), and could be matched to article identifiers in the Scopus bibliography. The process of selection and matching resulted in a dataset of 634,100 publications (Materials and Methods). We constructed article level clusters of the computer science literature for the years 1996-2015, at a sufficiently high level of aggregation to be comparable to disciplinary classification by journal. Clustering of publications is commonly accomplished through direct citation, bibliographic coupling, and co-citation with an argument being made for direct citation as the best approach to concentrate citation links~\cite{klavans_which_2017}. Accordingly, we first used direct citation links as the basis for cluster formation. A set of 20 clusters was generated using a spectral clustering technique, Graclus, that has previously been applied to citation data~\cite{graclus_2007,subelj_clustering_2016}. To place these clusters in context, we identified ASJC categories that accounted for at least 10\% of the publications in each cluster and observed 31 of the 334 minor subject areas in the Scopus ASJC classification. (which consists of 5 subject areas, 27 major subject areas, and 334 minor subject areas).

@@

Here we describe the overall strategy used for the two clustering methods.\\
